---
title: "Mixed Model Workshop part 3"
author: "Lorna Le Stanc"
date: "10/22/2021"
output:
  html_document:
    toc: yes
    number_sections: yes
    df_print: paged
    fig_width: 7
    fig_height: 4
  pdf_document:
    toc: yes
    number_sections: yes
    fig_width: 5
    fig_height: 4
---

# Goal of this workshop session

In this part three we will learn how to do post-hocs and planned contrasts for mixed models with categorical predictors.

A quick paragraph on mixed model assumptions. Really not exhautive. 

# Needed libraries

```{r, message = FALSE, warning=FALSE}
# to perform planned contrasts & post-hocs /!\ load before tidyverse to prevent 
# masking its 'select' function
library(multcomp)

# to manipulate and plot data
library(tidyverse)

# to do mixed models
library(lme4)

# to obtain main effects from linear models and linear mixed models
library(rstatix)

```

# Dataset for the workshop

Here we will work on reaction times (continuous variable)

-   Two groups (control vs test)

-   Two sessions (T0 vs T1)

-   4 items per sessions (each with 5 trials) -\> 20 trials per participant and session.

This is a 2x2 design, with group as a between-subjects variables and session as a within-subject variable.

The dataset was built such as :

```{r, eval = F}
# Number of subjects per groups (2 groups, control/test)
N = 30
# number of trials per subject
ntrials = 20
# reaction time at T0 for both groups (ms)
int.T0 = 950
# Retest effect
slope.retest = 100
# Training effect
slope.training = 150
# variance among subjects at T0
sd = 50
# variance among retest and training effects is sd/2
```

```{r, echo = FALSE, message=FALSE, eval = T}
data = read_csv("../data/dataset_groupXsessionXage.csv")%>%
  rename(participant = pax)%>%
  select(-age)

data.mean = data %>%
  group_by(participant, group,session)%>%
  summarise(rt.mean.bPbS = mean(rt))%>%
  ungroup()
```

```{r echo=FALSE, message=FALSE, fig.align = "center"}

ggplot(data.mean, aes(y = rt.mean.bPbS, x = session,fill = group))+
  geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 0.5, position = "dodge")+
  stat_summary(geom = "pointrange",
    fun = mean,
    fun.min = function(x) mean(x) - sd(x),
    fun.max = function(x) mean(x) + sd(x),
    col = "darkgoldenrod2",
    position = position_dodge(width = 0.9),
    show.legend = FALSE)+
  theme_bw()
```

# Summary of previous results

We arrived at the following model : 

```{r}
lme.conv = lmer(rt ~ group*session + (1+session|participant) + (1|item), data = data) 
```

With the following main effects : 

```{r}
Anova(lme.conv)
```

There is an interaction between group and session showing that the effect of session is different according to groups,  which we want to investigate. 


```{r}
summary(glht(lme.conv))
```

Given the output of the `lme.conv` model, we already have some information :

- We know that there is no difference between groups at T0 (grouptest). $\beta = 5.4 ± 13.4, z = 79.9, p=.97$

- We know that reaction times are faster at T1 for the control group (sessionT1). $\beta = -99.7 ± 6.9, z = -14.5, p<.001$

- We know that the effect of session is different between groups (grouptest:sessionT1) $\beta = -143.4 ± 9.7, z = -14.8, p<.001$

But we DONT have the following information : 

- Is there a difference between groups at T1 and what is the estimate (a ?) ?
- Is there a difference between sessions in the test group and what is the estimate (b ?) ?

That's why we want to do post-hocs.

```{r, echo = FALSE, message=FALSE}

data.interaction = data.mean%>%
  mutate(session = as.factor(session))%>%
  mutate(new_session = case_when(
    group == "control" ~ as.numeric(session)-0.225,
    group == "test" ~ as.numeric(session)+0.225
  ))%>%
  mutate(group = as.factor(group))%>%
  mutate(new_group = case_when(
    session == "T0" ~ as.numeric(group),
    session == "T1" ~ as.numeric(group)
  ))

```

```{r, echo = FALSE, message=FALSE}
lm = lm(rt.mean.bPbS ~ group*session, data = data.mean)

ggplot(data.interaction, aes(y = rt.mean.bPbS, x = session,fill = group))+
  geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 0.5, position = "dodge")+
  stat_summary(geom = "pointrange",
    fun = mean,
    fun.min = function(x) mean(x) - sd(x),
    fun.max = function(x) mean(x) + sd(x),
    col = "darkgoldenrod2",
    position = position_dodge(width = 0.9),
    show.legend = FALSE)+
  # stat_summary(fun = mean, geom = "line", col = "darkgoldenrod2", aes(x = new_session, group = group)) +
  theme_bw()+
  geom_text(x=0.5, y=lm$coefficients[[1]], label="Intercept")+
  geom_segment(aes(x = 0.8, y = lm$coefficients[[1]], xend = 1.2, yend = lm$coefficients[[1]] +lm$coefficients[[2]]), color = "darkgoldenrod2")+
  geom_text(x=1, y=lm$coefficients[[1]]+50, label="grouptest")+
  geom_segment(aes(x = 0.8, y = lm$coefficients[[1]], xend = 1.8, yend = lm$coefficients[[1]]+lm$coefficients[[3]]), color = "darkgoldenrod2")+
  geom_text(x=1, y=lm$coefficients[[1]]+lm$coefficients[[3]], label="sessionT1")+
  geom_segment(aes(x = 1.2, y = lm$coefficients[[1]]+lm$coefficients[[2]], xend = 2.2, yend = lm$coefficients[[1]]+lm$coefficients[[2]]+lm$coefficients[[3]]+lm$coefficients[[4]]), linetype = "dashed", color = "darkgoldenrod2")+
  geom_text(x=2, y=790, label="b ?")+
  geom_segment(aes(x = 1.8, y = lm$coefficients[[1]]+lm$coefficients[[3]], xend = 2.2, yend = lm$coefficients[[1]]+lm$coefficients[[2]]+lm$coefficients[[3]]+lm$coefficients[[4]]),  linetype = "dashed", color = "darkgoldenrod2")+
  geom_text(x=1.95, y=840, label="a ?")+
  theme_bw()+
  scale_x_discrete(breaks = c("T0", "T1"), labels = c("T0 = 0","T1 = 1"))+
  scale_fill_manual(breaks = c("control", "test"), labels = c("control = 0", "test = 1"), values = c("#F8766D", "#00BFC4"))

```

# Post-hocs

Here I will use the glht function from multcomp package to perform post-hoc analyses. 


```{r, echo=F}
data.post = data%>%
  mutate(group = as.factor(group),
         session = as.factor(session))
```

## Tukey comparison

Usually you see that on internet

```{r, warning=F}
summary(glht(lme.conv, linfct = mcp(group = "Tukey")),test=adjusted("none")) 
```

Warning, Tukey means that you compare all levels against each other. You test for all possible comparisons. But it is not a Tukey correction. Here there is no correction for multiple comparison `test=adjusted("none")`.

Here there are only two levels so one comparison. If you had 3 levels, you would have 2 comparisons etc. 

> Note that again, the results gives you the comparison at T0

## How to define a contrast ? OR Why I've been annoying you with math 

Another way to do post-hocs is to define the contrast you want to test. 

```{r}
# contrast
KgroupT0 = rbind(c(0,1,0,0))
# testing for the contrast, with no correction for multiple comparison : 
# we have only one comparison
summary(glht(lme.conv, linfct = KgroupT0),test=adjusted("none"))
```

Same result. 

You see what does not appear in usual outputs : the hypothesis tested is whether the contrast is different from 0. In R the equality test is written `==`.

### Write down the math of fixed effects for your model

Remember : 

---

<center> rt.mean.bPbS = Intercept + grouptest\*group + sessionT1\*session + grouptest:sessionT1\*group\*session 
</center>

Where

  - session = 0 if T0,  session =  1 if T1
  - group = 0 if control, group = 1 if test
  - Intercept is the mean in ref level (session = 0, group = 0)

---

WARNING : in the same order as in the model, you start building a table : 

|rt =                    |Intercept  |+ (grouptest) x group |+ (sessionT1) x session|+ (grouptest:sessionT1) x group x session|
|------------------------|----------:|---------------------:|----------------------:|----------------------------------------:|
|contrast                |           |                      |                       |                                         |


```{r}
summary(glht(lme.conv))
```

You always multiply your intercept by one. It is never null, always there in your equation : 

|rt =                    |Intercept  |+ (grouptest) x group |+ (sessionT1) x session|+ (grouptest:sessionT1) x group x session|
|------------------------|----------:|---------------------:|----------------------:|----------------------------------------:|
|contrast                |   1       |                      |                       |                                         |

For test group at T0, group = 1, session = 0 : 


|rt =                    |Intercept  |+ (grouptest) x group |+ (sessionT1) x session|+ (grouptest:sessionT1) x group x session|
|------------------------|----------:|---------------------:|----------------------:|--------------------------------------:|
|rt of tests at T0       | 1         | 1                    | 0                     | 1\*0 = 0                              |

For test group at T1, group = 1, session = 1 : 


|rt =                    |Intercept  |+ (grouptest) x group |+ (sessionT1) x session|+ (grouptest:sessionT1) x group x session|
|------------------------|----------:|---------------------:|----------------------:|--------------------------------------:|
|rt of tests at T0       | 1         | 1                    | 0                     | 1\*0 = 0                              |
|rt of tests at T1       | 1         | 1                    | 1                     | 1\*1 = 1                              |

> Testing the effect of session in the test group mathematically means testing whether (rt of tests at T1 - rt of tests at T0) is different from 0

|rt =                    |Intercept  |+ (grouptest) x group |+ (sessionT1) x session|+ (grouptest:sessionT1) x group x session|
|------------------------|----------:|---------------------:|----------------------:|--------------------------------------:|
|rt of tests at T0       | 1         | 1                    | 0                     | 1\*0 = 0                              |
|rt of tests at T1       | 1         | 1                    | 1                     | 1\*1 = 1                              |
|T1 vs T0 for tests      | 0         | 0                    | 1                     | 1                                     |

The last line is (line 2 - line 1). 

Be careful with the direction of your sustraction. It will not change the stats but the sign of the estimate. 

Mathematically : 

---
<center>
T1 vs T0 for tests = rt of tests at T1 - rt of tests at T0 \
rt of tests at T1 - rt of tests at T0 > 0 **means** rt of tests at T1 > rt of tests at T0\
rt of tests at T1 - rt of tests at T0 < 0 **means** rt of tests at T1 < rt of tests at T0\
</center>

---

If you define T1 vs T0 for tests = rt of tests at T0 - rt of tests at T1, a positive estimate will mean longer reaction time at T1 and a negative estimate will mean shorter reaction times at T1. It is the exact opposite meaning than in the previous definition.  

### Translate into R code

```{r}
# define the contrast
KsessionTests = rbind(c(0,0,1,1))
# give it a name to keep track (this will appear in the output of the test)
rownames(KsessionTests) = "Session Tests"
# test it (no correction)
summary(glht(lme.conv, linfct = KsessionTests),test=adjusted("none"))
```

The effect of session is significant in the test group. Reaction times are shorter at T1 by 243 ms in this group. 

We can verify this result by releveling the group factor in the model to define the test group as the reference. 

*sessionT1* will be the effect of session in the reference group ie the test group : 

```{r}

lme.group.relevel = lmer(rt ~ relevel(group, ref = "test") * session + 
                              (1 + session | participant) + (1 | item), 
                         data = data.post)
lme.group.relevel

```

### Correct for multiple comparisons

As we have done for the session effect in the test group, we can define the contrast for the effect of session in the control group : 

|rt =                    |Intercept  |+ (grouptest) x group |+ (sessionT1) x session|+ (grouptest:sessionT1) x group x session|
|------------------------|----------:|---------------------:|----------------------:|--------------------------------------:|
|rt of controls at T1    | 1         | 0                    | 1                     | 0\*1 = 0                              |
|rt of controls at T0    | 1         | 0                    | 0                     | 0\*0 = 0                              |
|T1 vs T0 for controls   | 0         | 0                    | 1                     | 0                                     |


```{r}
# define the contrast for the sontrol groups
KsessionControls = rbind(c(0,0,1,0))
# give it a name to keep track
rownames(KsessionControls) = "Session Controls"
# combine the two contrasts
Ksession = rbind(KsessionControls,KsessionTests)
# test it and control for multiple comparison with bonferroni correction
summary(glht(lme.conv, linfct = Ksession),test=adjusted("bonferroni"))
```

I've chosen to test the effects of session in the two groups. I could have tested the effect of group in the different sessions : 


|rt =                    |Intercept  |+ (grouptest) x group |+ (sessionT1) x session|+ (grouptest:sessionT1) x group x session|
|------------------------|----------:|---------------------:|----------------------:|--------------------------------------:|
|rt of controls at T0    | 1         | 0                    | 0                     | 0\*0 = 0                              |
|rt of test at T0        | 1         | 1                    | 0                     | 1\*0 = 0                              |
|tests vs controls at T0 | 0         | 1                    | 0                     | 0                                     |
|------------------------|-----------|----------------------|-----------------------|---------------------------------------|
|rt of controls at T1    | 1         | 0                    | 1                     | 0\*1 = 0                              |
|rt of test at T1        | 1         | 1                    | 1                     | 1\*1 = 1                              |
|tests vs controls at T1 | 0         | 1                    | 0                     | 1                                     |


```{r}
# contrast for T1
KgroupT1 = rbind(c(0,1,0,1))
# contrast for T0
KgroupT0 = rbind(c(0,1,0,0))
# combine contrasts
Kgroup = rbind(KgroupT0,KgroupT1)
# give names to keep track
rownames(Kgroup) = c("group at T0","group at T1")
# test while correcting for multiple comparisons
summary(glht(lme.conv, linfct = Kgroup),test=adjusted("bonferroni"))

```

Again we can check by releveling the session this time : 

```{r}
lme.session.relevel = lmer(rt ~ group * relevel(session, ref = "T1") 
                                + (1 + session | participant) + (1 | item), 
                           data = data.post)
lme.session.relevel
```

# Planned contrasts

As we defined the post-hocs, we can also define other contrasts. 

## Main effects

Lets define the main effect of session : 

Mathematically : 

---
<center>
T1 vs T0 = mean rt at T1 across groups - mean rt of at T0 across groups \
mean rt at T1 across groups = (rt at T1 for tests + rt at T1 for controls) /2\
mean rt at T0 across groups = (rt at T0 for tests + rt at T0 for controls) /2
</center>

---

Let's figure out the contrasts : 


|rt =                        |Intercept    |+ (grouptest) x group |+ (sessionT1) x session|+ (grouptest:sessionT1) x group x session|
|----------------------------|------------:|---------------------:|----------------------:|--------------------------------------:|
|rt at T1 for tests          | 1           | 1                    | 1                     | 1\*1 = 1                              |
|rt at T1 for controls       | 1           | 0                    | 1                     | 0\*1 = 0                              |
|mean rt at T1 across groups | (1+1)/2 = 1 | (1+0)/2 = 0.5        |(1+1)/2 = 1            | (1+0)/2 = 0.5                         |
|----------------------------|-------------|----------------------|-----------------------|---------------------------------------|
|rt at T0 for tests          | 1           | 1                    | 0                     | 1\*0 = 0                              |
|rt at T0 for controls       | 1           | 0                    | 0                     | 0\*0 = 0                              |
|mean rt at T0 across groups | (1+1)/2 = 1 | (1+0)/2 = 0.5        |(0+0)/2 = 0            | (0+0)/2 = 0                           |
|----------------------------|-------------|----------------------|-----------------------|---------------------------------------|
|T1 vs T0                    | 1-1 = 0     | 0.5-0.5 = 0          | 1-0 = 1               | 0.5-0 = 0.5                           |
|----------------------------|-------------|----------------------|-----------------------|---------------------------------------|


```{r}
Ksessionmain=rbind(c(0,0,1,0.5))
rownames(Ksessionmain) = "Main effect of session"
summary(glht(lme.conv, linfct = Ksessionmain),test=adjusted("none"))
```


In the same way you can define the main effect of group : 


|rt =                        |Intercept    |+ (grouptest) x group |+ (sessionT1) x session|+ (grouptest:sessionT1) x group x session|
|------------------------|----------:|---------------------:|----------------------:|--------------------------------------:|
|mean rt for controls    | 1         | 0                    | 0.5                   | 0\*0.5 = 0                            |
|mean rt for tests       | 1         | 1                    | 0.5                   | 1\*0.5 = 0.5                          |
|tests vs controls       | 0         | 1                    | 0                     | 0.5                                   |



```{r}
Kgroupmain=rbind(c(1,1,0.5,0.5)-c(1,0,0.5,0))
rownames(Kgroupmain) = "Main effect of group"
summary(glht(lme.conv, linfct = Kgroupmain),test=adjusted("none"))
```

## Interactions

In the same way you can define the contrast for the interaction.

Mathematically : 

---
<center>
interaction = effect of session for controls - effect of session for tests \
</center>
---

From previously defined contrast we get : 

|rt =                        |Intercept    |+ (grouptest) x group |+ (sessionT1) x session|+ (grouptest:sessionT1) x group x session|
|------------------------|----------:|---------------------:|----------------------:|--------------------------------------:|
|T1 vs T0 for controls   | 0         | 0                    | 1                     | 0                                     |
|T1 vs T0 for tests      | 0         | 0                    | 1                     | 1                                     |
|interaction             | 0         | 0                    | 0                     | 1                                     |

```{r}

Kinteraction=KsessionTests-KsessionControls # same as KgroupT1-KgroupT0
rownames(Kinteraction) = "Interaction"
Kinteraction
summary(glht(lme.conv, linfct = Kinteraction),test=adjusted("none"))

```

## Planned contrast

```{r}
# plan your contrast and test all at once
Kultimate=rbind(Kgroupmain,
                Ksessionmain,
                Kinteraction,
                Ksession)
summary(glht(lme.conv, linfct = Kultimate),test=adjusted("bonferroni"))
summary(lme.conv)
```


# Mixed Model Assumptions

## Hypothesis of linearity (not necesssary when categorical predictors)

## normality of residuals

```{r}
qqnorm(resid(lme.conv))
qqline(resid(lme.conv), col = "red")

lme.log = lmer(log(rt) ~ group * session + (1 + session | participant) + (1 | item), data = data)

qqnorm(resid(lme.log))
qqline(resid(lme.log), col = "red")
```

```{r, echo = FALSE, eval = T}
summary(lme.log)
summary(lme.conv)
```

```{r}
plot(lme.conv)
```


## normality of random intercept, random slopes

```{r}
qqnorm(ranef(lme.conv)$participant[,1] )
qqline(ranef(lme.conv)$participant[,1], col = "red")

qqnorm(ranef(lme.conv)$participant[,2] )
qqline(ranef(lme.conv)$participant[,2], col = "red")

qqnorm(ranef(lme.conv)$item[,1] )
qqline(ranef(lme.conv)$item[,1], col = "red")
```


```{r, eval = FALSE, echo = F}
lme.c = lmer(rt ~ session*age + (1+session|participant), data = data)
summary(lme.c)
```

